---
title: "Importation and representation of parabiosis droplet data"
author: 
    - name: Tram Nguyen
      affiliation: Center for Computational Biomedicine, Harvard Medical School
      email: Tram_Nguyen@hms.harvard.edu
    - name: Kris Holton
      affiliation: Harvard Stem Cell Institute, Harvard Medical School
    - name: Tyrone Lee
      affiliation: Center for Computational Biomedicine, Harvard Medical School
    - name: Nitesh Turaga
      affiliation: Center for Computational Biomedicine, Harvard Medical School
    - name: Ludwig Geistlinger
      affiliation: Center for Computational Biomedicine, Harvard Medical School
    - name: Robert Gentleman
      affiliation: Center for Computational Biomedicine, Harvard Medical School
package: MouseAgingData
output:
    BiocStyle::html_document:
      self_contained: yes 
      toc: true
      toc_float: true
      toc_depth: 2
      code_folding: show
date: "`r doc_date()`"
vignette: >
    %\VignetteIndexEntry{parabiosis droplet}
    %\VignetteEncoding{UTF-8}
    %\VignetteEngine{knitr::rmarkdown}
editor_options: 
    markdown: 
      wrap: 80
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    crop = NULL
)
```

# Installation

Install the package using Bioconductor. Start R and enter:

```{r, eval = FALSE}
# if(!requireNamespace("BiocManager", quietly = TRUE))
    # install.packages("BiocManager")
# BiocManager::install("MouseAgingData")
```

# Setup

Now, load the package and dependencies used in the vignette.

```{r, message = FALSE}
library(scran)
library(scater)
library(ggplot2)
library(bluster)
library(SingleCellExperiment)
library(ExperimentHub)
library(MouseAgingData)
```

# Introduction

Single-cell sequencing technology can reveal intricate details about individual
cells, allowing researchers to interrogate the genetic make up of cells within a
heterogeneous sample. Single-cell sequencing can provide insights into various
aspects of cellular biology, such as characterization of cell populations,
identification of rare cell types, and quantification of expression levels in
cell types across experimental treatments. Given the wide utility, single-cell
sequencing has expanded scientific knowledge in various fields, including cancer
research, immunology, developmental biology, neurobiology, and microbiology.

There are several methods for generating single-cell sequencing data which can
extract information (DNA or RNA) from a cell. These include, but are not limited
to:

1. Droplet-based platforms: such as 10x Genomics Chromium system, inDrop,
Drop-seq, and Seq-Well, which use microfluidic devices to isolate individual
cells into tiny droplets along with unique barcoded beads.

2. Plate or microwell-based methods: such as the Smart-seq2 protocol or the C1
system by Fluidigm, respectively. These platforms employ microfluidic chips or
multi-well arrays to capture and process individual cells. Unlike droplet-based
platforms, these cells are manually or automatically sorted into individual
wells of the plate.

The `MouseAgingData` package provides analysis-ready data from an aging mouse
brain parabiosis single cell study by Ximerakis & Holton et al.,
([2023](https://pubmed.ncbi.nlm.nih.gov/37118429/)) and additional datasets. 
The contents of the package can be accessed by querying ExperimentHub with the 
package name.


# Data

Ximerakis & Holton et al. investigated how heterochronic parabiosis (joining of
the circulatory systems) affects the mouse brain in terms of aging and
rejuvenation. They identified gene signatures attributed to aging in specific
cell-types. They focus especially on brain endothelial cells, which showed
dynamic transcriptional changes that affect vascular structure and function.

The parabiosis single cell RNA-seq (Ximerakis, Holton et al Nature Aging 2023)
includes 105,329 cells, 31 cell types across 8 OX, 8 YX, 7 YY, 9 YO, 7 OO, 11 OY
animals, and 20905 features.

This vignette performs a simple analysis of the parabiosis 10x Genomics
single-cell data set, following the Quick Start Workflow of Single-Cell Analysis
in the [OSCA Bioconductor Book](https://bioconductor.org/books/3.15/OSCA.intro/analysis-overview.html#quick-start-simple). 

Briefly, it walks through the process of quality control, normalization, various
forms of dimensionality reduction, clustering, detection of marker genes, and
annotation of cell types. PCA, UMAP, and tSNE coordinates used in the study were
provided by the authors for visualization.


# Load the data set from ExperimentHub

```{r}
sce <- parabiosis10x()
```

```{r}
# View the data
sce
```

Do some checking to make sure the data loaded correctly and is what we expected.

```{r, Data check}
# Sample metadata
head(colData(sce)) 
```


```{r}
# Includes cell colors from the original paper
metadata(sce)
```


# Quality control

In this step, we can explore and visualize mitochondrial content and read count.
However, the authors have already removed low-quality cells and animals so we
will skip this section in this vignette. For more details on their workflow, one
can refer to the original article Ximerakis & Holton et al.
([2023](https://www.nature.com/articles/s43587-023-00373-6)). The [OSCA
Bioconductor
book](https://bioconductor.org/books/3.15/OSCA.basic/quality-control.html) also
provides several examples of quality control steps as well.


# Normalization

Normalize the expression counts. For the purposes of demonstration, we'll subset
this `SingleCellExperiment` object down to the first 1000 cells.

```{r, subset sce}
sce_subset <- sce[, 1:1000]
```

```{r, normalize}
set.seed(101000110)
clusters <- quickCluster(sce_subset)
sce_subset <- computeSumFactors(sce_subset, clusters=clusters)
sce_subset <- logNormCounts(sce_subset)

logcounts(sce_subset)[1:10, 1:10]
```

# Feature selection

At this point in a typical workflow, we could select an appropriate set of
highly variable genes (HVGs), say the top 10% of genes with the highest
variability in expression. Below is an example of how to do this with our
subsetted `SingleCellExperiment` example.

```{r, feat select}
dec <- modelGeneVar(sce_subset)
hvg <- getTopHVGs(dec, prop=0.1)
```

<br>

However, a logical index showing the 2000 HVG included in the original study
conducted by the authors can also be accessed in the original
`SingleCellExperiment` object in the `rowData()` slot.

```{r}
head(rowData(sce))
```


# PCA

Below is a method for running a Principal Components Analysis using our
previously defined HVGs. Since this step can take a significant amount of time
to compute, we will again just apply it to our subset of 1000 cells as
demonstration.

```{r, PCA, warning=FALSE}
# Since we already have PCA coords from our authors, we will add these computed 
# PCA coords under a different name "osca_PCA"

set.seed(1234)
sce_subset <- runPCA(sce_subset, ncomponents=25, subset_row=hvg, 
                     name = "osca_PCA")


# Show the names of the elements in the ReducedDims slot
reducedDims(sce_subset)
```
<br>

As mentioned, the authors have also provided us with the first 50 PCs used in
their study within the full `SingleCellExperiment` object. Let's take a look.

```{r, viewPCA, warning=FALSE}
reducedDim(sce_subset, "PCA")[1:5, 1:5]
```

# Clustering

At this point, we could take the PCs that were previously computed and do some
clustering of cells based on expression profiles. More details are provided in
the OSCA book
[here](https://bioconductor.org/books/3.14/OSCA.basic/clustering.html). Let's do
some clustering with our subsetted object as an example.

```{r, clustering}
colLabels(sce_subset) <- clusterCells(sce_subset, use.dimred='osca_PCA',
                               BLUSPARAM=NNGraphParam(cluster.fun="louvain"))
```


# Visualization

For this dataset, the authors have already provided us with their exact UMAP and
tSNE coordinates, as well as their color scheme representing the cell types from
their paper. This can be accessed in the metadata slot of the
`SingleCellExperiment` object with the `metadata()` function. To consistently
recreate their figures, let's plot using their provided coordinates.

```{r, fig.wide=TRUE}

# Generate color map matching cell type to colors in publication
cell.color <- metadata(sce)$cell_color

gg <- plotUMAP(sce, color_by = "cell_type", text_by = "cell_type") 
gg + theme(legend.title=element_blank()) + 
    scale_color_manual(values=c(cell.color))
```
This plot is a recreation of Fig. 2C from Ximerakis & Holton et al. 2023.

<br>

We can also plot a tSNE with their provided coordinates.

```{r, plot provided tSNE, fig.wide=TRUE}
gg <- plotTSNE(sce, color_by = "cell_type", text_by = "cell_type") 
gg + theme(legend.title=element_blank()) + 
    scale_color_manual(values=c(cell.color))
```

<br>

If you would like to create your own UMAP and tSNE plots, please refer back to
the [OSCA Bioconductor book](https://bioconductor.org/books/3.15/OSCA.basic/)
for more details.

# Reference

Ximerakis & Holton et al. (2023) Heterochronic parabiosis reprograms the mouse
brain transcriptome by shifting aging signatures in multiple cell types.
\emph{Nat Aging} 3, 327–345. <DOI:https://doi.org/10.1038/s43587-023-00373-6>.

# Session Info

```{r, sesh info}
sessionInfo()
```
