---
title: "MouseAgingData Intro"
author: "Tram N Nguyen"
package: MouseAgingData
output: 
    BiocStyle::html_document
vignette: >
    %\VignetteIndexEntry{MouseAgingData Intro}
    %\VignetteEncoding{UTF-8}
    %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 80
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

# Load Mouse Aging Data

The `MouseAgingData` package provides analysis-ready data resources of different
aging omics datasets, including the aging mouse brain single cell study
(Ximerakis et al., 2019), aging mouse brain parabiosis single cell study
(Ximerakis et al., 2023). In addition, two relevant transcriptomic studies of
rejuvenation in the GEO database are also available, including: an aging mouse
caloric restriction single cell study (GSE137869, Ma et al., 2020), and an aging
mouse exercise single cell study (GSE196364, Liu et al., 2022). The contents of
the package can be accesssed by querying the ExperimentHub with the package
name.

```{r}
suppressPackageStartupMessages({
    library(ExperimentHub)
    library(SingleCellExperiment)
    library(MouseAgingData)
})

eh <- ExperimentHub()
query(eh, "MouseAgingData")
```

The individual data sets can be accessed using either their ExperimentHub
accession number, or the provided functions in this package. For example, to
load the parabiosis 10X Genomics droplet data from Ximerakis et al. 2023, do the
following:

```{r}
droplet <- eh[["EH1617"]]
droplet
droplet <- parabiosis10X()
droplet
```

To view a simple workflow to analyze and visualize the downloaded data following
the [OSCA Single-Cell
Workbook](https://bioconductor.org/books/3.18/OSCA.basic/), do the following:

```{r}
browseVignettes("MouseAgingData")
```

# Session info

```{r}
sessionInfo()
```
