---
title: "MouseAgingData Intro"
author: "Tram N Nguyen"
package: MouseAgingData
output: 
    BiocStyle::html_document
vignette: >
    %\VignetteIndexEntry{MouseAgingData Intro}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

# Load Mouse Aging Data

The `TabulaMurisData` data package provides access to the 10x and SmartSeq2
single-cell RNA-seq data sets from 
[the Tabula Muris Consortium](http://tabula-muris.ds.czbiohub.org/). 
The contents of the package can be seen by querying the ExperimentHub for the
package name.

```{r}
suppressPackageStartupMessages({
    library(ExperimentHub)
    library(SingleCellExperiment)
    library(MouseAgingData)
})

eh <- ExperimentHub()
query(eh, "MouseAgingData")
```

The individual data sets can be accessed using either their ExperimentHub
accession number, or the convenience functions provided in this package. For
example, for the 10x data:

```{r}
droplet <- eh[["EH1617"]]
droplet
droplet <- parabiosis10X()
droplet
```


# Session info

```{r}
sessionInfo()
```

