---
title: "Import and representation of Mouse Aging Data"
author: 
    - name: Tram Nguyen
      affiliation: Center for Computational Biomedicine, Harvard Medical School
      email: Tram_Nguyen@hms.harvard.edu
    - name: Kris Holton
      affiliation: Harvard Stem Cell Institute, Harvard Medical School
    - name: Tyrone Lee
      affiliation: Center for Computational Biomedicine, Harvard Medical School
    - name: Ludwig Geistlinger
      affiliation: Center for Computational Biomedicine, Harvard Medical School
    - name: Robert Gentleman
      affiliation: Center for Computational Biomedicine, Harvard Medical School
package: MouseAgingData
output:
    BiocStyle::html_document:
      self_contained: yes 
      toc: true
      toc_float: true
      toc_depth: 2
      code_folding: show
date: "`r doc_date()`"
vignette: >
    % \VignetteIndexEntry{Mouse ileum}
    %\VignetteEncoding{UTF-8}
    % \VignetteEngine{knitr::rmarkdown}
editor_options: 
    markdown: 
      wrap: 80
---

```{r style, echo = FALSE, results = 'asis', include = FALSE}
  BiocStyle::markdown()
```

# Installation

Install the package using Bioconductor. Start R and enter:

```{r, eval = FALSE}
if(!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("MouseAgingData")
```

# Setup

Now, load the package and dependencies used in the vignette.

```{r, setup, message = FALSE}
library(DropletUtils)
library(edgeR)
library(Seurat)
library(scran)
library(ggplot2)
library(SingleCellExperiment)
library(ExperimentHub)
```

# Data

The `MouseAgingData` package provides analysis-ready data resources of different
aging omics datasets, including the aging mouse brain single cell study
Ximerakis et al., [2019](https://www.nature.com/articles/s41593-019-0491-3),
aging mouse brain parabiosis single cell study Ximerakis & Holton et al.,
[2023](https://pubmed.ncbi.nlm.nih.gov/37118429/). In addition, two relevant
transcriptomic studies of rejuvenation are also provided: an aging mouse caloric
restriction single cell study Ma et al.,
[2020](https://pubmed.ncbi.nlm.nih.gov/32109414/), and an aging mouse exercise
single cell study Liu et al., [2022](https://pubmed.ncbi.nlm.nih.gov/37080206/).
The contents of the package can be accessed by querying ExperimentHub with the
package name.

This vignette performs a simple analysis of the parabiosis 10X Genomics
single-cell dataset from Ximerakis & Holton et al.
([2023](https://www.nature.com/articles/s43587-023-00373-6)). The analysis
follows the Basics of Single-Cell Analysis in the [OSCA Bioconductor book Quick
Start
Workflow](https://bioconductor.org/books/3.15/OSCA.intro/analysis-overview.html#quick-start-simple).

Steps in this vignette include: quality control, normalization, various forms of
dimensionality reduction, clustering into subpopulations, detection of marker
genes, and annotation of cell types.

```{r}
eh <- ExperimentHub()
query(eh, c("MouseAgingData", "parabiosis10X"))
```

### About the dataset

Parabiosis single cell RNA-seq (Ximerakis, Holton et al Nature Aging 2023)
105,329 cells, 31 cell types 8 OX, 8 YX, 7 YY, 9 YO, 7 OO, 11 OY 20905 features

Processed: (ALL) reduced to: OLG, OPC, GLUT, GABA, EC, ASC Aging OX v YX; RJV:
OY v OX; AGA YO v YX

<br>

### Make the `SingleCellExperiment` object.

Load cell barcodes, genes, and matrix files using "DropletUtils".

```{r, Data Loading}
library(MouseAgingData)

# getWebData(format = "tsv", destination = "~/Desktop/", url = "https://ftp.ncbi.nlm.nih.gov/geo/series/GSE222nnn/GSE222510/suppl/GSE222510_processed_barcodes.tsv.gz", header=F)

# Eventually we can load in data through our package, for now:
sce <- readRDS("~/Desktop/parabiosis_sce.rds")
sce
```

Do some quick QC to make sure the data loaded correctly and is what we expected.

```{r, Data check}
# Check sce dimensions
dim(sce)

# Sample metadata
head(colData(sce)) 

# Includes cell colors from the original paper
metadata(sce)

# The count matrix. Transcript for each gene per cell (column)
head(counts(sce)) 
```

## 3. Quality control

In this step, we can explore and visualize mitochondrial content and read count.
However, the authors have already removed low-quality cells and animals so we
will skip this section in this vignette. For more details on their workflow, one
can refer to the original article Ximerakis et al.
([2023](https://www.nature.com/articles/s43587-023-00373-6)). The [OSCA
Bioconductor
book](https://bioconductor.org/books/3.15/OSCA.basic/quality-control.html) also
provides several examples of quality control steps as well.

## 4. Normalization

```{r, normalize}
library(scran)
set.seed(101000110)
clusters <- quickCluster(sce)
sce <- computeSumFactors(sce, clusters=clusters)
sce <- logNormCounts(sce)


head(logcounts(sce))
```

## 5. Feature selection

Select an appropriate set of highly variable genes (HVGs).

```{r, feat select}
library(scran)
dec <- modelGeneVar(sce)
hvg <- getTopHVGs(dec, prop=0.1)
```

## 6. PCA

```{r, PCA, warning=F}
library(scater)
set.seed(1234)
sce <- runPCA(sce, ncomponents=25, subset_row=hvg) # Add PCA coords to sce object
```

## 7. Clustering

```{r, clustering}
library(bluster)
colLabels(sce) <- clusterCells(sce, use.dimred='PCA',
                               BLUSPARAM=NNGraphParam(cluster.fun="louvain"))
```

## 8. Visualization

For this dataset, the authors have already provided us with their exact UMAP and
TSNE coordinates, as well as their color scheme representing the cell types from
their paper. This can be accessed in the metadata slot of the
`SingleCellExperiment` object with the `metadata()` function.

```{r provided umap, fig.height = 4, fig.width = 6.5}
# Generate color map matching cell type to colors in publication
cell.color <- metadata(sce)$cell_color

library(scater)
gg <- plotUMAP(sce, color_by = "cell_type", text_by = "cell_type") 
gg + theme(legend.title=element_blank()) + scale_color_manual(values=c(cell.color))
```

This plot is a recreation of Fig. 2C from Ximerakis et al. 2023.

We can also plot a TSNE with their provided coordinates.

```{r, plot provided TSNE, fig.height = 4, fig.width = 6.5}
gg <- plotTSNE(sce, color_by = "cell_type", text_by = "cell_type") 
gg + theme(legend.title=element_blank()) + scale_color_manual(values=c(cell.color))
```

<br>

We can also create our own UMAP and TSNE plots following the OSCA workflow and
save our own coordinates to the `SingleCellExperiment` object.

Start with a UMAP.

```{r, UMAP2, fig.height = 4, fig.width = 6.5}
# Add our UMAP to the reducedDims slot with the name "osca_UMAP" so we do not confused it with the author's UMAP coords.
sce <- runUMAP(sce, dimred = 'PCA', name = "osca_UMAP")

# When we want to access a specific item (here, "osca_UMAP") in the reducedDims element of the SingleCellExperiment object, we can use the function plotReducedDim with the argument dimred.

gg <- plotReducedDim(sce, dimred = "osca_UMAP", color_by = "cell_type", text_by = "cell_type") 
gg + theme(legend.title=element_blank()) + scale_color_manual(values=c(cell.color))
```

Now with a TSNE.

```{r, TSNE2, fig.height = 4, fig.width = 6.5}
# Run TSNE
set.seed(00101001101)

# runTSNE() stores the t-SNE coordinates in the reducedDims
sce <- runTSNE(sce, dimred = 'PCA', name = "osca_TSNE")

gg <- plotReducedDim(sce, dimred = "osca_TSNE", color_by = "cell_type", text_by = "cell_type") 
gg + theme(legend.title=element_blank()) + scale_color_manual(values=c(cell.color))
```

## Reference

Ximerakis et al. (2023) Heterochronic parabiosis reprograms the mouse brain
transcriptome by shifting aging signatures in multiple cell types.
\emph{Nat Aging} 3, 327–345. <DOI:https://doi.org/10.1038/s43587-023-00373-6>.

## View Session Info

```{r, sesh info}
sessionInfo()
```
