---
title: "Preparing Ximerakis 2019NN brain 10x droplet dataset for ExperimentHub"
author: "Tram Nguyen"
date: "2024-03-01"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(width = 80)
```

# Data pre-processing

Basic processing of the 10x Genomics scRNA-seq data was already
performed with the Seurat package (v.2.3) in R (v.3.3.4) to remove
low-quality cells and identify cell clusters following the detailed
methods described in Ximerakis et al.
[(2019)](https://www.nature.com/articles/s41593-019-0491-3#Sec12).

Briefly, the data were log normalized and scaled to 10,000 transcripts
per cell and highly variable genes were identified with the
FindVariableGenes() function. Next, principal component analysis (PCA)
was carried out, and the top 20 principal components (PCs) were stored.
Clusters were identified with the FindClusters() function with the
shared nearest neighbor method. All clusters with only one cell were
removed, resulting in 40 initial clusters.

For initial quality-control (QC) filtering, the authors removed any
clusters where the majority of cells had greater than 30%
mitochondrial RNA, under 1,000 detected transcripts, or under 500 unique
genes. Additionally, they filtered the remaining individual cells with
the following parameters: minimum percentage mito=0, maximum
percentage mito=30%, minimum number of UMI=200, maximum number of
UMIs=30,000, minimum number of nGene=250, and maximum number of
nGene=6,000 to exclude outliers. Finally, any genes that were
detected in fewer than 3 cells were excluded. After initial QC, a total of 
38,244 cells and 14,699 genes were retained.

A final round of QC was performed to remove doublet artifacts arising from
the co-capture of multiple cells in one droplet. Using the top differential
markers, the authors searched for identified cluster/sub-cluster using the
FindMarkers() function. Doublets/multiplets were defined as any cluster in
which >30% of its cells were expressed in at least 5 of the top 10 genes
specific for the initially identified cell type and also any other cell type it 
was associated with. These clusters were removed from downstream analysis.
Furthermore, cell clusters that were not represented by at least half of the
young and old animals were also excluded. After exclusions, clustering was again
performed.

The final dataset includes the transcriptomes derived from the brains of
8 young (2–3 months) and 8 old (21–23 months) mice, and includes 37,089 cells
representing 25 cell types and 14699 features. 


# Download processed data

The raw, unprocessed data for the 16 mice samples are available for download 
through GEO with accession number [GSE129788](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE12978).

Additional cell metadata and UMAP, TSNE, and PCA coordinates were provided 
courtesy of the authors.

# Generate `SingleCellExperiment` object
```{r}
# Convert Seurat to SingleCellExperiment object
s_sce<-as.SingleCellExperiment(s2, assay="RNA")

# Format cell metadata
s_sce$orig.ident<-NULL
cols <- c("animal", "batch", "animal_type", "cell_type")
colData(s_sce)[cols] <- lapply(colData(s_sce)[cols], factor)

# Add cell type colors matching paper
metadata(s_sce)$cell_colors<-s2@misc$colours

# Format reduced dim elements
reducedDims(s_sce) <- reducedDims(s_sce)[c("PCA","TSNE")]
colnames(reducedDim(s_sce, "TSNE")) <- c("TSNE1", "TSNE2")
colnames(reducedDim(s_sce, "PCA")) <-  sub("_", "", colnames(reducedDim(s_sce, "PCA")))

# Format row and column data
colnames(s_sce) <- NULL
rowData(s_sce) <- NULL

# Retain only raw counts
s_sce@assays@data@listData$logcounts <- NULL
s_sce@assays@data@listData$scaledata <- NULL
```
